\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-05-27 Latex/main.tex
updates

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{bbold}
\usepackage{float}
\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

So far, the studies about narwhals behavioral responses to sound exposure prove that they can exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns \cite{heide-jorgensen_behavioral_2021}. Moreover, a significant decrease in their buzzing rate has been assessed as far as 40 km away from the sound source \cite{tervo_narwhals_2021}. These results are based on controlled exposure experiments conducted in 2017 and 2018 in South-east Greeland, which is home to a declining population of narwhals \cite{garde_biological_2022}.  Though GPS positions are retrieved only when the narwhals emerge to the surface, that is irregularly in time, they have so far been analyzed with discrete time methods relying on linear interpolation to obtain data every second. In \cite{heide-jorgensen_behavioral_2021}, the positions are classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second, and the dynamic of the states is modeled with a Markov chain, assuming that the current state only depends on the last one in time. An exposure covariate defined as the inverse of the distance to the sound source is then used to model the transition probabilities of the Markov Chain. The probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases. Additionally, an overall speed increase due to sound exposure is revealed by an analysis of the consecutive GPS positions through a mixed effect model. 
Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one single model. \\

We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations represents an extension of the CVM which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. Here, $W(t)$ is a two-dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}. Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study. Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions. We follow \cite{gurarie_correlated_2017} and refer to~\ref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\

@@

Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}. To make this possible, we derive explicit formulas for the transitioon density of the process $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}$ and then formulate a state-space model from the discretisation of these formulas. Maximum likelihood estimation based on Kalman filter is performed as for any other model in \texttt{smoothSDE}.\\

@@

In Section 2, we give an overview of the narwhal data available for the analysis.
Then, the diffusion models are discussed in Section 3. Section 4 is dedicated to the statistical model used to infer the shore and sound exposure effect. Simulation study results are shown in section 5. Eventually, our models are applied to analyse a behavioral response of the narwhals in section 6.

@@

The dataset analysed in this section has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters. The Scoresby Sound fjord system
is known to be the summer residence for an isolated population of narwhals.
Here we recall briefly how the data has been retrieved. For more details about the study area and the tagging of the animals, the reader may refer to \cite{heide-jorgensen_behavioral_2021}. 
An offshore patrol vessel military ship was sailed to shoot airguns underwater between August 25 and September 1. It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots. The guns were fired synchronously every 80 s, and the GPS navigation system recorded the location of every shot. 
Data includes narwhals's latitude and longitude, distance relative to the ship in metres, GPS time, distance to the shore in metres. GPS positions are known only at specific times when the narwhals get close to the sea surface. The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours. While statistical analysis in \cite{heide-jorgensen_behavioral_2021} relied on positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.

@@

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$. Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary. About $9\%$ of the GPS measurements in the dataset turned out to be on land. After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy the of land polygons. We modified  the land polygons using \texttt{qgis}. After adjustements,  approximately $1 \%$ of the observations remained on land. Distance to shore values range from $0$ to $7.6$ km. We define an exposure to the shore covariate with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
		E^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=830$ m, which is the median of observed distances. Thus we only consider the effect of the shore when the narwhals are close enough to it. 
The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is estimated as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard.png}
		\caption{}
	\end{subfigure}
	\caption{(a) Trajectory of a RCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}. (b) Trajectory of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}. Position are sampled from constant time step $\Delta= 1$ min and then downsampled to $5$ min. The samples are obtained from approximate transition densities  with constant covariates on each time step.}
	\label{fig: sample RACVM}
\end{figure}

\subsection{Constrained rotational correlated velocity model}
\label{section: CRCVM}

Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
To get a constrained version of the CVM, we propose to make use of a new RCVM that includes how the animals turn in reaction to the shore. To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
We keep the notation $\mathcal{D} \subset \R^2$ for the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

@@

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
In this case, the formulas of section~\ref{section: RACVM} are still valid.
The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be close to $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show a smooth function obtained from equation~\ref{eq: smooth omega}, while figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM). A sample of a CRCVM with smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}. The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system in South-East Greenland.

@@

 We show side by side the samples obtained with a circular border for a constant value of $\tau$ and for a smooth function $\tau$ defined as in figure~\ref{fig:smoothtaubump}.

 \begin{figure}[H]
 	\centering
 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.4]{images/crcvm/illustrative_sample_circ_standard}
 		\label{fig:illustrativesamplecircstandard}
 	\end{subfigure}
 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.4]{images/crcvm/illustrative_sample_circ_pers}
 		\label{fig:illustrativesamplecircpers}
 	\end{subfigure}

 	\caption{Left: sample from a CRCVM with smooth $\omega$ and constant $\tau=1$ h.  Right: sample from a CRCVM with smooth $\omega$ and smooth $\tau$ to represent persistent movement along the boundary.}
 	\label{fig: crcvm examples}
 \end{figure}

@@

\section{Statistical model}

\subsection{Mixed effet model for ship exposure and shore effects}
\label{subsection: ship exposure effect}

@@

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}\\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}  \\
		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t)) \quad t \geq 0\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact baseline model}
\end{equation}

@@

\begin{equation}  \forall j \in \{1,\cdots,n_{i,post}\}, \quad \left\{
	\begin{array}{l}
		\log(\tau_{ij})=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k})\phi_{k}(E^{ship}_i(t))+b_{\tau,i} \\
		\log(\nu_{ij})=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k}) \phi_{k}(E^{ship}_i(t)) +b_{\nu,i}  \\
	\end{array}
	\right.
	\label{eq: exact response model}
\end{equation}

Again, the functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.

>>>>

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}

\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{bbold}

\usepackage[margin=0.8in]{geometry}
\usepackage[backend=biber,style=authoryear,backref = true]{biblatex}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{xcolor}
\usepackage{nicematrix}
\usepackage{multirow}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{yhmath}
\usetikzlibrary{quotes,angles}
\usepackage{pgfplots}
\pgfplotsset{compat = newest}
\pagestyle{fancy}
\fancyhf{}
\fancyhfoffset[L]{1cm} 
\fancyhfoffset[R]{1cm} 
 \setlength{\headheight}{14pt}
\rhead{Greenland narwhal's behavioural responses}
\lhead{\bfseries \today}
\cfoot{\thepage}

@@

So far, the studies about narwhals behavioral responses to sound exposure prove that they can exhibit avoidance reactions and have a proclivity to move toward the shore when exposed to seismic airguns \cite{heide-jorgensen_behavioral_2021}. Moreover, a significant decrease in their buzzing rate has been assessed as far as 40 km away from the sound source \cite{tervo_narwhals_2021}. These results are based on controlled exposure experiments conducted in 2017 and 2018 in South-east Greeland, which is home to a declining population of narwhals \cite{garde_biological_2022}.  Though GPS positions are retrieved only when the narwhals emerge to the surface, that is irregularly in time, they have so far been analyzed with discrete time methods relying on linear interpolation to obtain data every second. In \cite{heide-jorgensen_behavioral_2021}, the positions are classified in three states "Far from the shore", "Moving toward shore" and "Close to the shore" at each second, and the dynamic of the states is modeled with a Markov chain, assuming that the current state only depends on the last one in time. An exposure covariate defined as the inverse of the distance to the sound source is then used to model the transition probabilities of the Markov Chain. The probabilities of being in states "Close to the shore" and "Moving toward the shore" are shown to increase with exposure to the sound, while probability of being in state "Far from the shore" decreases. Additionally, an overall speed increase due to sound exposure is revealed by an analysis of the consecutive GPS positions through a mixed effect model. 
Here, we  make use of a continuous time model to analyse the GPS positions in order to be able to quantify the effect of sound exposure on the speed and the heading of the narwhals in one holistic model. \\

We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations represents an extension of the CVM which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. Here, $W(t)$ is a two-dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}. Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study. Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source. We follow \cite{gurarie_correlated_2017} and refer to~\ref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\

@@

Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}. To make this possible, we derive explicit formulas for the transition density of the process $U(t)=\begin{pmatrix} X(t) & V(t) \end{pmatrix}$ and then formulate a state-space model from the discretisation of these formulas. Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\

@@

In Section 2, we give an overview of the narwhal data available for the analysis.
Then, the diffusion models are discussed in Section 3. Section 4 is dedicated to the statistical model used to infer the shore and sound exposure effects. Simulation study results are shown in section 5. Eventually, our models are applied to analyse a behavioral response of the narwhals in section 6.

@@

The dataset analysed in this paper has been the subject of several studies, focusing mainly on vocalizations and avoidance reactions \cite{heide-jorgensen_behavioral_2021,tervo_narwhals_2021}. 
Six male narwhals were equipped with FastLoc GPS receivers in August 2018 in Scoresby Sound fjord system in South-East Greenland by biologists from the Greenland Institute of natural ressources, with the help of local Inuit hunters. The Scoresby Sound fjord system
is known to be the summer residence for an isolated population of narwhals.
Here we recall briefly how the data has been retrieved. For more details about the study area and the tagging of the animals, the reader may refer to \cite{heide-jorgensen_behavioral_2021}. 
An offshore patrol vessel military ship was sailed to shoot airguns underwater between August 25 and September 1. It was equipped with two Sercel G-guns at 6m depth and moved at a speed of 4.5 knots. The guns were fired synchronously every 80 s, and the GPS navigation system recorded the location of every shot. 
Data includes narwhals's latitude and longitude, distance relative to the ship in metres, GPS time, distance to the shore in metres. GPS positions are known only at specific times when the narwhals get close to the sea surface. The median time step between two GPS measurements in the data is about $5$ minutes and only $0.3 \%$ of the time steps reach more than two hours, with a maximum at more than $4$ hours. While statistical analysis in \cite{heide-jorgensen_behavioral_2021} relied on positions linearly interpolated at each second, here we only consider the actual GPS measurements with irregular time steps.

@@

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$. Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary. About $9\%$ of the GPS measurements in the dataset turned out to be on land. After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy the of land polygons. After adjustement the land polygons using \texttt{qgis},  approximately $1 \%$ of the observations remained on land. Distance to shore values range from $0$ to $7.6$ km. We define an exposure to the shore covariate with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
		E^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=830$ m, which is the median of observed distances. Thus we only consider the effect of the shore when the narwhals are close enough to it. 
The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is estimated as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/rcvm/illustrative_sample_standard.png}
	\caption{Trajectory of a RCVM with parameters $\tau=1$h, $\nu=4$ km/h, $\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.}
	\label{fig: sample RACVM}
\end{figure}

Unconstrained movement of marine mammals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$, while other animals or organisms whose motion appear tortuous require the more general model \ref{eq: RACVM equation}.
In the case of motion under spatial constraints, an increase in tortuosity enables the animals to avoid the boundary and adapt their heading to the constraint. 
\subsection{Constrained rotational correlated velocity model}
\label{section: CRCVM}

We thus propose to make use of a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to include how the animals turn in reaction to the shore. 
To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
We keep the notation $\mathcal{D} \subset \R^2$ for the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

@@

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
In this case, we can use the formulas of section~\ref{section: RACVM} to get an approximation of the exact transition density.
The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be close to $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated  :

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show a smooth function obtained from equation~\ref{eq: smooth omega}, while figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).

@@

 We illustrate this flexibility in Figure~\ref{fig: crcvm examples}.Trajectories obtained within the fjords domain for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.

 \begin{figure}[H]

 	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.65]{images/crcvm/illustrative_sample_fjords}
		\label{fig:illustrativesamplefjords}
	\end{subfigure}

 	\caption{Five day trajectories with time step $\Delta=5$ min in Scoresby Sound fjords domain}
 	\label{fig: crcvm examples}
 \end{figure}

@@

\section{Inference of the sound exposure effect}

\subsection{Mixed effet model for sound exposure and shore effects}
\label{subsection: ship exposure effect}

@@

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}\\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}  \\
		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
		\omega_i(t)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E_i^{shore},\Theta_i)
	\end{array}
	\right.
	\label{eq: exact baseline model}
\end{equation}

@@

\begin{equation}   \left\{
	\begin{array}{l}
		y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots,n_{i,post}\} \\
		\log(\tau_{i}(t))=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k})\phi_{k}(E^{ship}_i(t))+b_{\tau,i} \\
		\log(\nu_{i}(t))=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k}) \phi_{k}(E^{ship}_i(t)) +b_{\nu,i}  \\
	\end{array}
	\right.
	\label{eq: exact response model}
\end{equation}

The intercepts $\log(\tau_0)$, $\log(\nu_0)$ and the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are considered are offset.
The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate. Only the parameters $\log(\tau_k)$ and $\log(\nu_k)$ are estimated from the trajectories during exposure to the sound
